\SW: II. New Gravitational Lens Candidates from the CFHTLS Discovered through Citizen Science

### 2015-05-14 doc/sw-cfhtls.tex
fixed some numbers in Table 2, corresp. and other minor edits

\begin{abstract}
We report the discovery of 29 promising (and 59 total) new lens
candidates from the CFHT Legacy Survey (\cfhtls) based on about 11
million classifications performed by citizen scientists as part of the
first \sw lens search.  The goal of the blind lens search was to
identify lens candidates missed by robots (the \rf on galaxy scales and
\af on group/cluster scales) which had been previously used to mine the
\cfhtls for lenses.  We compare some properties of the samples detected
by these algorithms to the \sw sample and find them to be broadly
similar.  The image separation distribution calculated from the \sw
sample shows that previous constraints on the average density profile of
lens galaxies are robust. \sw recovers about 60\% of known lenses, while
the new candidates show a richer variety compared to those found by the
two robots. This detection rate could be increased to 75\% by only using
classifications performed by expert volunteers (albeit at the cost of a
higher false positive rate), indicating that the training and
performance calibration of the citizen scientists is very important for
the success of \sw. In this work we present the SIMCT pipeline, used for
generating {\it in situ} a sample of realistic  simulated lensed images.
This training sample, along with the false positives identified during
the search, has a legacy value for testing future lens finding
algorithms. We make the pipeline and the training set publicly available.
\end{abstract}

@@

In order to calculate the lensing cross-section, we first calculate the
luminosity of each potential lensing galaxy using the photometric redshifts
($z_l$) from the parent galaxy catalog. Next, we use the $L-\sigma$ scaling relation from the
bright sample of \citep{Parker2005} given by
\be
\label{magstar2}
\sigma_v=142 \left(\frac{L}{L_{*}}\right)^{1/3} \, Km~s^{-1}.
\ee
This sets the velocity dispersion of the halo
hosting the galaxy, which will be later used in the model.
We assume that the knee of the luminosity function of galaxies, $L_*$ evolves
such that there is a decline of $1.5$ magnitudes between $z=1$ to $z=0$
\citep{Faber2007}.\footnote{We anchor our $L_*$ evolution at low redshifts using the
determination of $L_*$ in the r-band by \citet{Blanton2001}. To maintain
consistency in magnitude systems, we have converted the CFHT MegaCAM magnitudes
to SDSS magnitudes and k-corrected them to $z=0.1$.}

@@

\begin{table}
\begin{center}
\caption{ \label{tab:stats}
Statistics of detections in \sw }
\begin{tabular}{l l l l l}
\hline
   &  \multicolumn{2}{c} {\StageOne}  & \multicolumn{2}{c}{\StageTwo} \\
      & KC  &  KL  & KC & KL \\
\hline
\hline
Number  & 128 & 34 & 107  & 34  \\
Fraction  & & & & \\
of total recovered & 29\% & 58\% & 25\% & 58\%  \\
$P_{\rm acc thresh}$ & 0.95 & 0.95 & 0.3 & 0.3 \\
Averaged Grade (G) & $-$ & $-$ & 1.3 & 1.3 \\
\hline
   &  \multicolumn{2}{c} {\StageTwo}  &   & \\
      & NC  &  AC  &  & \\
\hline
\hline
Number  & 59 & 141 &  & \\
Fraction & & & & \\
of detections & 14\% & 33\% &  & \\
$P_{\rm acc thresh}$ & 0.3 & 0.3 &  & \\
Averaged Grade (G) & 1.3 & 1.3 & 1.3 & 1.3 \\

@@

In \Tref{tab:stats}, we give overall statistics of the systems detected
at \StageOne and \StageTwo. We give the total number of detections of
the known lens candidates, known lenses and the new lens candidates at
each stage.  We also give the recovery fractions for the known samples
and fraction of total detection for the new samples at each stage. We
find that all of the confirmed lenses found at \StageOne are also
recovered at \StageTwo; a small fraction of the known lens candidates
are rejected at \StageTwo. From the visually inspected known lens sample
at \StageTwo (P$> 0.3$), the experts gave a grade of 1.3 and above to
$\sim$~92~per~cent (33 out of 36) of the systems. Overall, the sample of
new \sw lens candidates increased the sample of total \cfhtls lens
candidates by over 40~per~cent. However, we find that 42~per~cent of
known lenses were missed at \StageOne: we return to the discussion of
these false negatives below.

In \Fref{fig:stackP}, we plot the distribution of false positives and
the high grade lens candidates,
as a function of the $P$ value assigned by SWAP at the end of \StageTwo.
On average, the fraction of lens candidates is indeed an increasing
function of $P$. This shows that the \sw generated $P$ values for the
subject are roughly correlated with the expert grades albeit with quite
some scatter.  We note that below $P\sim0.75$,
the fraction of false positives starts to exceed the fraction of real
lens candidates. This could be a good threshold to choose to maximize
the purity of the final sample. However, choosing
$P=0.3$ gives a completeness of 100\% for the ``known
lens'' sample instead of 67\% for $P=0.75$. Therefore, the new sample
should also have increased completeness; expert
grading then allows us to increase the purity of the sample.

@@

We now determine the fraction of the known sample of lenses that were recovered
by \sw. Note that this sample corresponds to the \rf and \af samples
combined, as defined in \Sref{sec:data:kls}. In \Tref{tab:stats},
we show that $\sim25-30\%$ of the known lens candidates, and
$\sim60\%$ of the known lenses were found at \StageOne. We find that
$84$\% of the known lens candidates and \textbf{all} of the known lenses
from \StageOne passed our \StageTwo selection criterion of $P>0.3$.
The left and the middle panels of \Fref{fig:compre} show the
fraction of detections as a function of arc magnitude and the Einstein
radius of the lens systems for the known confirmed lenses and lens
candidates. As expected, we find that systems with brighter images
and/or with larger $R_{\rm E}$ are detected more often in \sw.

@@

In subsequent \rf steps, the flux from the galaxy is subtracted from the
scaled difference image to enhance the visibility of the faint blue
lensed features. An object finder is then run on this image to quantify the
lensed image properties.
Another $\sim$~50~per~cent of the \sw candidates could not be
detected by the object finder because properties such as the image area,
axis ratio, magnitude/colour and alignment with respect to the lensing
galaxy were not satisfied. Some of the candidates missed at this stage
are, for example, SW4, SW5, SW6, SW26, SW36, SW39 and SW46.

@@

Like any lens finding method, the \sw system could potentially be failing
to detect certain
kinds of lenses.  We find that about 40~per~cent of the known sample of lenses
are missed at \StageOne; no further losses were incurred during
the \StageTwo refinement (see \Tref{tab:stats}).
Below, we focus on the known
lens sample at \StageOne to understand why some of them are being missed
and possibly find a way to improve the detection rate which can be
adopted in the future \sw lens searches.

Many of the missed lenses are from the \rf sample with small
Einstein radii and faint lensed images (see \Fref{fig:compre}). Among
the confirmed lenses from \rf, about 50~per~cent are missed. Out of the
missed sample of 18 lenses, about half of them are visually difficult to
detect. The other half appear to have faint blue smudges around
galaxies which should have been easier to identify. Similarly,
if we consider the \af lens sample, $\sim$23~per~cent are missed by \sw. This
is a relatively small sample of $\sim 5$ systems and visual inspection
of them suggests that, by and large, either the lensed features are faint
or they have odd properties which makes them difficult to identify
correctly. For further tests, we combine the \rf and \af sample.

@@

\item We use a sample of simulated lenses, duds and impostors tailored
to the \cfhtls data to train the volunteers and calibrate their
performance. The volunteers not only perform well on the training sample
(see Paper I) but also find lenses that are fainter, more compact or
redder than covered by the training sample demonstrating their
adaptability in this task.

@@

\item Based on the known sample of lenses and lens candidates, we find
that we lose a very small fraction of them during \StageTwo refinement.
It is more important to improve the lens
detection sensitivity at the initial \StageOne classification step.
About 40\% of the previously known lenses ($\sim$20 in total) were missed at \StageOne.
Roughly half of these missed lenses were found to be galaxy-scale \rf
systems, with faint arcs blended with the bright lens galaxies.

\item It is possible to improve the \sw completeness by
changing the strategy of when and who is shown an image: only using high power
classifications recovers 30\% of the missed lenses.

@@

The discovery of many new lens candidates through the first \sw lens
search has demonstrated that the citizen scientists have successfully
taught themselves to identify lenses within a short span of time.  They
have found lens candidates which the algorithms failed to discover.
Upcoming and planned wide-field imaging surveys such as the DES, HSC and
LSST will produce formidable amounts of data.  Blind lens searches as
described here will be infeasible with these very large surveys although
it may be possible to do a blind search on a subset of a large survey in
order to assess the performance of either the algorithms or the
volunteers which can be extrapolated to the entire survey. However, any
one approach for finding lenses from the entire survey data is likely
going to be inefficient. Thus, combining robotic methods for
pre-selection with the citizen science approach for visual screening
could be a good strategy for finding lenses in these large imaging
surveys.

>>>>

\begin{abstract}
We report the discovery of 29 promising (and 59 total) new lens
candidates from the CFHT Legacy Survey (\cfhtls) based on about 11
million classifications performed by citizen scientists as part of the
first \sw lens search.  The goal of the blind lens search was to
identify lens candidates missed by robots (the \rf on galaxy scales and
\af on group/cluster scales) which had been previously used to mine the
\cfhtls for lenses.  We compare some properties of the samples detected
by these algorithms to the \sw sample and find them to be broadly
similar.  The image separation distribution calculated from the \sw
sample shows that previous constraints on the average density profile of
lens galaxies are robust. \new{ \sw recovers about 65\% of known lenses, while
the new candidates show a richer variety compared to those found by the
two robots. This detection rate could be increased to 80\% by only using}
classifications performed by expert volunteers (albeit at the cost of a
higher false positive rate), indicating that the training and
performance calibration of the citizen scientists is very important for
the success of \sw. In this work we present the SIMCT pipeline, used for
generating {\it in situ} a sample of realistic  simulated lensed images.
This training sample, along with the false positives identified during
the search, has a legacy value for testing future lens finding
algorithms. We make the pipeline and the training set publicly available.
\end{abstract}

@@

In order to calculate the lensing cross-section, we first calculate the
luminosity of each potential lensing galaxy using the photometric redshifts
($z_l$) from the parent galaxy catalog. Next, we use the $L-\sigma$ scaling relation from the
bright sample of \citep{Parker2005} given by
\be
\label{magstar2}
\sigma_v=142 \left(\frac{L}{L_{*}}\right)^{1/3} \, {\rm Km~s^{-1}}.
\ee
This sets the velocity dispersion of the halo
hosting the galaxy, which will be later used in the model.
We assume that the knee of the luminosity function of galaxies, $L_*$ evolves
such that there is a decline of $1.5$ magnitudes between $z=1$ to $z=0$
\citep{Faber2007}.\footnote{We anchor our $L_*$ evolution at low redshifts using the
determination of $L_*$ in the r-band by \citet{Blanton2001}. To maintain
consistency in magnitude systems, we have converted the CFHT MegaCAM magnitudes
to SDSS magnitudes and k-corrected them to $z=0.1$.}

@@

\begin{table}
\begin{center}
\caption{ \label{tab:stats}
\new {Statistics of detections in \sw} }
\begin{tabular}{l l l l l}
\hline
   &  \multicolumn{2}{c} {\StageOne}  & \multicolumn{2}{c}{\StageTwo} \\
      & KC  &  KL  & KC & KL \\
\hline
\hline
Number  & 142 & 39 & 79  & 34  \\
Fraction  & & & & \\
of total recovered & 32\% & 66\% & 18\% & 58\% \\
$P_{\rm acc thresh}$ & 0.95 & 0.95 & 0.3 & 0.3 \\
Averaged Grade (G) & $-$ & $-$ & 1.3 & 1.3 \\
\hline
   &  \multicolumn{2}{c} {\StageTwo}  &   & \\
      & NC  &  AC  &  & \\
\hline
\hline
Number  & 59 & 141 &  & \\
Fraction & & & & \\
of detections & 14\% & 33\% &  & \\
$P_{\rm acc thresh}$ & 0.3 & 0.3 &  & \\
Averaged Grade (G) & 1.3 & 1.3 & &  \\

@@

\new{In \Tref{tab:stats}, we give overall statistics of the systems
detected at \StageOne and \StageTwo. We give the total number of
detections of the known lens candidates, known lenses and the new lens
candidates at each stage.  We also give the recovery fractions for the
known samples and fraction of total detection for the new samples at
each stage.  Overall, the sample of new \sw lens candidates comprises
over 40~per~cent of the total \sw$-$\cfhtls lens candidates.  Nearly
45~per~cent of the known lens candidates from \StageOne are rejected at
\StageTwo. Such fractions are acceptable for ``candidates'' as their
quality grades vary from high to low.  We find that most of the
confirmed lenses found at \StageOne are also recovered at \StageTwo.
However, $\sim35$~per~cent of the known lenses were missed already at
\StageOne: we return to the discussion of these false negatives below in
\Sref{sec:fn}.
}

In \Fref{fig:stackP}, we plot the distribution of false positives and
the high grade lens candidates, as a function of the $P$ value assigned
by SWAP at the end of \StageTwo.  On average, the fraction of lens
candidates is indeed an increasing function of $P$. This shows that the
\sw generated $P$ values for the subject are roughly correlated with the
expert grades albeit with quite some scatter.  We note that below
$P\sim0.75$, the fraction of false positives starts to exceed the
fraction of real lens candidates. This could be a good threshold to
choose to maximize \new{the purity of the final sample. However,
choosing $P=0.3$ gives a completeness of 92\% for the ``known lens''
sample instead of 64\% for $P=0.75$.} Therefore, the new sample should
also have increased completeness; expert grading then allows us to
increase the purity of the sample.

@@

We now determine the fraction of the known sample of lenses that were
recovered by \sw. Note that this sample corresponds to the \rf and \af
samples combined, as defined in \Sref{sec:data:kls}. In
\new{ \Tref{tab:stats}, we show that $\sim32\%$ of the known lens candidates,
and $\sim65\%$ of the known lenses were found at \StageOne. We find that
$56$\% of the known lens candidates and $87$\% of the known lenses from
\StageOne passed our \StageTwo selection criterion of $P>0.3$ and
$G>1.3$.} The left and the middle panels of \Fref{fig:compre} show the
fraction of detections as a function of arc magnitude and the Einstein
radius of the lens systems for the known confirmed lenses and lens
candidates. As expected, we find that systems with brighter images
and/or with larger $R_{\rm E}$ are detected more often in \sw.

@@

In subsequent \rf steps, the flux from the galaxy is subtracted from the
scaled difference image to enhance the visibility of the faint blue
lensed features. An object finder is then run on this image to quantify the
lensed image properties.

Another $\sim$~50~per~cent of the \sw candidates could not be
detected by the object finder because properties such as the image area,
axis ratio, magnitude/colour and alignment with respect to the lensing
galaxy were not satisfied. Some of the candidates missed at this stage
are, for example, SW4, SW5, SW6, SW26, SW36, SW39 and SW46.

@@

\new{Like any lens finding method, the \sw system could potentially be
failing to detect certain kinds of lenses.  We find that about
35~per~cent of the known sample of lenses are missed at \StageOne; 
about 10~per~cent losses were incurred during the \StageTwo refinement
(see \Tref{tab:stats}).  Below, we focus on the known lens sample at
\StageOne to understand why some of them are being missed and possibly
find a way to improve the detection rate which can be adopted in the
future \sw lens searches.}

\new{Many of the missed lenses are from the \rf sample with small
Einstein radii and faint lensed images (see \Fref{fig:compre}). Among
the confirmed lenses from \rf, about 45~per~cent are missed. Out of the
missed sample of 15 lenses, about half of them are visually difficult to
detect. The other half appear to have faint blue smudges around
galaxies which should have been easier to identify. Similarly,
if we consider the \af lens sample, $\sim$20~per~cent are missed by \sw. This
is a relatively small sample of $\sim 5$ systems and visual inspection
of them suggests that, by and large, either the lensed features are faint
or they have odd properties which makes them difficult to identify
correctly. For further tests, we combine the \rf and \af sample.}

@@

\new {\item We use a sample of simulated lenses, duds and impostors tailored
to the \cfhtls data to train the volunteers and calibrate their
performance. The volunteers not only perform well on the training sample
(see Paper I) but also find lenses that are fainter, more compact or
redder than covered by the training sample demonstrating their
adaptability in this task. }

@@

\new{\item Based on the known sample of lenses and lens candidates, we find
that we lose a small fraction of them during \StageTwo refinement.
It is more important to improve the lens
detection sensitivity at the initial \StageOne classification step.
About 35\% of the known lenses (20 in total) were missed at \StageOne.
Two thirds of these missed lenses were found to be galaxy-scale \rf
systems, with faint arcs blended with the bright lens galaxies.}

\new{
\item It is possible to improve the \sw completeness by
changing the strategy of when and who is shown an image: only using high power
classifications recovers 40\% of the missed lenses.
}

@@

The discovery of many new lens candidates through the first \sw lens
search has demonstrated that the citizen scientists have successfully
taught themselves to identify lenses within a short span of time.  They
have found lens candidates which the algorithms failed to discover.
\new{Upcoming and planned wide-field imaging surveys such as the DES, HSC and
LSST will produce formidable amounts of data.  Blind lens searches as
described here will be infeasible with these very large surveys although
it may be possible to do a blind search on a subset of a large survey in
order to assess the performance of either the algorithms or the
volunteers which can be extrapolated to the entire survey. However, any
one approach for finding lenses from the entire survey data is likely
going to be inefficient.} Thus, combining robotic methods for
pre-selection with the citizen science approach for visual screening
could be a good strategy for finding lenses in these large imaging
surveys.
